Single-seismometer Focal Mechanism and Uncertainty Estimation from Body Waves

### 2024-10-29 paper/main.tex
Computational graph

% import packages
\usepackage{gensymb, amssymb, nicefrac, bm}

@@

    \begin{align} \label{eq:eq1}
      \nonumber &A^P \sim \nicefrac{\left( s_R(3\cos^2(i) - 1) - q_R\sin(2i) - p_R\sin^2(i)\right)}{\alpha_h^3}\\
      &A^{SV} \sim \nicefrac{\left( 1.5 s_R\sin(2j) + q_R\cos(2j) + 0.5 p_R\sin(2j) \right)}{\beta_h^3}\\
      \nonumber &A^{SH} \sim \nicefrac{\left( q_L\cos(j) + p_L\sin(j) \right)}{\beta_h^3}
    \end{align}

@@

    \begin{align} \label{eq:eq2}
      \nonumber s_R &= 0.5 \sin(\lambda) \sin(2\delta)\\
      \nonumber q_R &= \sin(\lambda) \cos(2\delta) \sin(\psi_r) + \cos(\lambda) \cos(\delta) \cos(\psi_r)\\
      p_R &= \cos(\lambda) \sin(\delta) \sin(2\psi_r) - 0.5\sin(\lambda) \sin(2\delta) \cos(2\psi_r)\\
      \nonumber p_L &= 0.5\sin(\lambda) \sin(2\delta) \sin(2\psi_r) + \cos(\lambda) \sin(\delta) \cos(2\psi_r)\\
      \nonumber q_L &= -\cos(\lambda) \cos(\delta) \sin(\psi_r) + \sin(\lambda) \cos(2\delta) \cos(\psi_r)
    \end{align}

@@

    \begin{align} \label{eq:eq3}
      \mu(\bm{p}) = \frac{\bm{A}_o^\top\bm{A}_s}{\|\bm{A}_o\|\|\bm{A}_s\|}
    \end{align}

@@

    \subsection{Justification}
    We are interested in situations where absolute amplitudes are either unknown or unnecessary, for
    instance if using data collected from NASA's Insight mission \citep{sita_potential_2022}.
    In this case, we cannot know the absolute velocities because they depend on an understanding of
    Mars' interior that is good enough to build a reliable velocity model. Alternative misfit
    functions can account for the relative amplitudes, such as in equations \ref{eq:eq4} and
    \ref{eq:eq5}:

    \begin{align} \label{eq:eq4}
      \mu(\bm{p}) = \sqrt{\left( \Delta \frac{A^{P}}{A^{SV}}\right)^2 + \left( \Delta \frac{A^{SV}}{A^{SH}}\right)^2}
    \end{align}

    \begin{align} \label{eq:eq5}
      \mu(\bm{p}) = \sqrt{\log^2 \left( \Delta \frac{A^P}{A^{SV}} \right) + \log^2 \left( \Delta \frac{A^{SV}}{A^{SH}} \right)}
    \end{align}

    First, we note that using cosine similarity is more numerically stable in cases where one of the
    observed or synthetic amplitudes is close to zero since the norm of the associated vector in
    equation \ref{eq:eq3} may still be large enough to make division feasible.

    Secondly, we have chosen cosine similarity over angle because it involves one less operation,
    which is more computationally efficient and easier to differentiate for directed search
    algorithms (see Section \ref{sec:algorithms}). The Gaussian function in equation \ref{eq:eq6} is
    similarly easy to differentiate, but it does not account for relative amplitudes so it would
    exclude otherwise acceptable solutions.

    \begin{align} \label{eq:eq6}
      &\mu(\bm{p}) \sim |\Sigma|^{-1/2} \exp \left( - \frac{1}{2} (\bm{A} - \bm{A}_o)^\top \bm{\Sigma}^{-1} (\bm{A} - \bm{A}_o) \right)
    \end{align}

    \subsection{Tolerance}
    Given vectors $\bm{A}_s$ and $\bm{A}_o$, we would like to know the largest cosine
    similarity for which $\bm{A}_s$ is considered an acceptable fit with respect to the recorded amplitudes
    while accounting for the asymmetry of error levels for different amplitudes. Visualized in 3D
    space, we have a confidence ellipsoid centered at $\bm{A}_o$ (equation \ref{eq:eq7}) such that
    acceptable fits are vectors which fall inside the elliptical cone tangent to it with vertex at the
    origin.
    \begin{align} \label{eq:eq7}
      E(\bm{p}) = (\bm{A} - \bm{A}_o)^\top \hat{\bm{\Sigma}}^{-1} (\bm{A} - \bm{A}_o)  = 1
    \end{align}

    \noindent where $\bm{A} = (A^P, A^{SV}, A^{SH})^\top$ is arbitrary
    and $\hat{\bm{\Sigma}} = \text{diag}(\sigma_P^2, \sigma_{SV}^2, \sigma_{SH}^2)/\chi^2_{2,\alpha}$ for some $\alpha \in (0,1)$. The elliptical
    cone intersects with the ellipsoid at every boundary point for which
    \begin{align} \label{eq:eq8}
      (\nabla_{\bm{A}} E)^\top \bm{A} = 2(\bm{A} - \bm{A}_o)^\top \hat{\bm{\Sigma}}^{-1} \bm{A} = 0
    \end{align}

    Combining equations \ref{eq:eq7} and \ref{eq:eq8} gives the plane containing every point of 
    tangency:
    \begin{align} \label{eq:eq9}
      \frac{1}{2} (\nabla_{\bm{A}} E)^\top \bm{A} - f(\bm{A}) = 
              \bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} (\bm{A} - \bm{A}_o)  = -1
    \end{align}

    To account for asymmetry, we recognize that the orientation of $\bm{A}_s$ relative to $\bm{A}_o$
    has an effect on the tolerance because the associated "thickness" of the ellipsoid changes.
    This information is captured by the plane containing vectors $\bm{A}_o$ and $\bm{A}_s$:
    \begin{align} \label{eq:eq10}
      (\bm{A}_s \times \bm{A}_o)^\top \bm{A} = 0
    \end{align}

    To find the constrained points of tangency, we can intersect the original ellipsoid with planes
    \ref{eq:eq9} and \ref{eq:eq10}. The line of intersection is such that
    \begin{align} \label{eq:eq11}
      \bm{A}(t) = (1 - \nicefrac{1}{c_1}) \bm{A}_o + t\bm{D}
    \end{align}

    \noindent where $c_1$ (equation \ref{eq:eq12}) is a location parameter projecting $\bm{A}_o$ onto plane \ref{eq:eq9}
    and $\bm{D}$ (equation \ref{eq:eq13}) is the direction perpendicular to planes \ref{eq:eq9} and \ref{eq:eq10}.

    \begin{align} \label{eq:eq12}
      \nonumber \bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} ((1 - \nicefrac{1}{c_1})\bm{A}_o - \bm{A}_o)  = -1\\
      \Rightarrow c_1 = \bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} \bm{A}_o = \|\bm{A}_o^\top, \bm{A}_o\|_{\hat{\bm{\Sigma}}^{-1}}^2
    \end{align}
    \begin{align} \label{eq:eq13}
      \nonumber\bm{D} &= (\bm{A}_o^\top \hat{\bm{\Sigma}}^{-1})^\top \times (\bm{A}_s \times \bm{A}_o)\\
        &= c_1\bm{A}_s - c_2\bm{A}_o
    \end{align}

@@

    Combining equations \ref{eq:eq7} and \ref{eq:eq11} produces a quadratic parameterized by the scalar $t$,
    from which we can obtain a quadratic equation for the points of tangency:

    \begin{align} \label{eq:eq14}
      \nonumber (t\bm{D} - \nicefrac{1}{c_1}\bm{A}_o)^\top \hat{\bm{\Sigma}}^{-1} (t\bm{D} - \nicefrac{1}{c_1}\bm{A}_o)  = 1\\
      \nonumber ({\bm{D}^\top \hat{\bm{\Sigma}}^{-1} \bm{D}})t^2 - \nicefrac{2}{c_1}(\bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} \bm{D})t + (\nicefrac{1}{c_1} - 1) = 0\\
      % \Rightarrow t = \nicefrac{c_3}{c_1c_2} \pm \nicefrac{1}{c_1c_2} \sqrt{c_3^2 + c_1c_2(c_1 - 1)}
    \end{align}

    Denote $c_3 = \bm{A}_s^\top \hat{\bm{\Sigma}}^{-1} \bm{A}_s = \|\bm{A}_s^\top, \bm{A}_s\|_{\hat{\bm{\Sigma}}^{-1}}^2$. Then by
    substitution we have
    ${\bm{D}^\top \hat{\bm{\Sigma}}^{-1} \bm{D}} = c_1(c_1c_3 - c_2^2)$ and ${\bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} \bm{D}} = 0$.
    Solving then yields the necessary points (denoted $\bm{A}_b$) as
    \begin{align} \label{eq:eq15}
      \bm{A}_b = \left(1 - \frac{1}{c_1}\right)\bm{A}_o \pm
        \left(\frac{1}{c_1} \sqrt{\frac{c_1 - 1}{c_1c_3 - c_2^2}}\right)(c_1\bm{A}_s - c_2\bm{A}_o)
    \end{align}

    Equation \ref{eq:eq15} shows that for the points to be well-defined, the following conditions
    must hold relative to the inner-product space defined by $\hat{\bm{\Sigma}}^{-1}$:

@@

    Thus, given 2 nonparallel vectors $\bm{A}_o$ and $\bm{A}_s$ and positive definite covariance
    matrix $\hat{\bm{\Sigma}}$, we accept the simulation if $\hat{\bm{A}}_o^\top \hat{\bm{A}_s} \geq \text{tol} 
    = \min_b \hat{\bm{A}}_o^\top \hat{\bm{A}_b}$. The condition, more formally, is
    \begin{align} \label{eq:eq16}
      \max_b \hat{\bm{A}}_o^\top (\hat{\bm{A}}_b - \hat{\bm{A}}_s) \leq 0
    \end{align}

@@

    The vector $\bm{A} = (A^P, A^{SV}, A^{SH})^\top$ can be treated as a continuously differentiable nonlinear function of
    parameters $\bm{p} = (\psi, \delta, \lambda)^\top$, defined by equation \ref{eq:eq1}. We can thus apply
    directed search methods to minimize the following cost function:
    \begin{align} \label{eq:eq17}
      \mu(\bm{p}) = \hat{\bm{A}}_o^\top \hat{\bm{A}}
    \end{align}

    TODO: draw computational graph, derive gradients

>>>>

% import packages
\usepackage{gensymb, amssymb, nicefrac, bm, tikz}

@@

    \begin{align} \label{eq:1}
      \nonumber &A^P \sim \nicefrac{\left( s_R(3\cos^2(i) - 1) - q_R\sin(2i) - p_R\sin^2(i)\right)}{\alpha_h^3}\\
      &A^{SV} \sim \nicefrac{\left( 1.5 s_R\sin(2j) + q_R\cos(2j) + 0.5 p_R\sin(2j) \right)}{\beta_h^3}\\
      \nonumber &A^{SH} \sim \nicefrac{\left( q_L\cos(j) + p_L\sin(j) \right)}{\beta_h^3}
    \end{align}

@@

    \begin{align} \label{eq:2}
      \nonumber s_R &= 0.5 \sin(\lambda) \sin(2\delta)\\
      \nonumber q_R &= \sin(\lambda) \cos(2\delta) \sin(\psi_r) + \cos(\lambda) \cos(\delta) \cos(\psi_r)\\
      p_R &= \cos(\lambda) \sin(\delta) \sin(2\psi_r) - 0.5\sin(\lambda) \sin(2\delta) \cos(2\psi_r)\\
      \nonumber p_L &= 0.5\sin(\lambda) \sin(2\delta) \sin(2\psi_r) + \cos(\lambda) \sin(\delta) \cos(2\psi_r)\\
      \nonumber q_L &= -\cos(\lambda) \cos(\delta) \sin(\psi_r) + \sin(\lambda) \cos(2\delta) \cos(\psi_r)
    \end{align}

@@

    \begin{align} \label{eq:3}
      \mu(\bm{p}) = \frac{\bm{A}_o^\top\bm{A}_s}{\|\bm{A}_o\|\|\bm{A}_s\|}
    \end{align}

@@

    \subsection{Justification}
    We are interested in situations where absolute amplitudes are either unknown or unnecessary, for
    instance if using data collected from NASA's Insight mission \citep{sita_potential_2022}.
    In this case, we cannot know the absolute velocities because they depend on an understanding of
    Mars' interior that is good enough to build a reliable velocity model. Alternative misfit
    functions can account for the relative amplitudes, such as in equations \ref{eq:4} and
    \ref{eq:5}:

    \begin{align} \label{eq:4}
      \mu(\bm{p}) = \sqrt{\left( \Delta \frac{A^{P}}{A^{SV}}\right)^2 + \left( \Delta \frac{A^{SV}}{A^{SH}}\right)^2}
    \end{align}

    \begin{align} \label{eq:5}
      \mu(\bm{p}) = \sqrt{\log^2 \left( \Delta \frac{A^P}{A^{SV}} \right) + \log^2 \left( \Delta \frac{A^{SV}}{A^{SH}} \right)}
    \end{align}

    First, we note that using cosine similarity is more numerically stable in cases where one of the
    observed or synthetic amplitudes is close to zero since the norm of the associated vector in
    equation \ref{eq:3} may still be large enough to make division feasible.

    Secondly, we have chosen cosine similarity over angle because it involves one less operation,
    which is more computationally efficient and easier to differentiate for directed search
    algorithms (see Section \ref{sec:algorithms}). The Gaussian function in equation \ref{eq:6} is
    similarly easy to differentiate, but it does not account for relative amplitudes so it would
    exclude otherwise acceptable solutions.

    \begin{align} \label{eq:6}
      &\mu(\bm{p}) \sim |\Sigma|^{-1/2} \exp \left( - \frac{1}{2} (\bm{A} - \bm{A}_o)^\top \bm{\Sigma}^{-1} (\bm{A} - \bm{A}_o) \right)
    \end{align}

    \subsection{Tolerance}
    Given vectors $\bm{A}_s$ and $\bm{A}_o$, we would like to know the largest cosine
    similarity for which $\bm{A}_s$ is considered an acceptable fit with respect to the recorded amplitudes
    while accounting for the asymmetry of error levels for different amplitudes. Visualized in 3D
    space, we have a confidence ellipsoid centered at $\bm{A}_o$ (equation \ref{eq:7}) such that
    acceptable fits are vectors which fall inside the elliptical cone tangent to it with vertex at the
    origin.
    \begin{align} \label{eq:7}
      E(\bm{p}) = (\bm{A} - \bm{A}_o)^\top \hat{\bm{\Sigma}}^{-1} (\bm{A} - \bm{A}_o)  = 1
    \end{align}

    \noindent where $\bm{A} = (A^P, A^{SV}, A^{SH})^\top$ is arbitrary
    and $\hat{\bm{\Sigma}} = \text{diag}(\sigma_P^2, \sigma_{SV}^2, \sigma_{SH}^2)/\chi^2_{2,\alpha}$ for some $\alpha \in (0,1)$. The elliptical
    cone intersects with the ellipsoid at every boundary point for which
    \begin{align} \label{eq:8}
      (\nabla_{\bm{A}} E)^\top \bm{A} = 2(\bm{A} - \bm{A}_o)^\top \hat{\bm{\Sigma}}^{-1} \bm{A} = 0
    \end{align}

    Combining equations \ref{eq:7} and \ref{eq:8} gives the plane containing every point of 
    tangency:
    \begin{align} \label{eq:9}
      \frac{1}{2} (\nabla_{\bm{A}} E)^\top \bm{A} - f(\bm{A}) = 
              \bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} (\bm{A} - \bm{A}_o)  = -1
    \end{align}

    To account for asymmetry, we recognize that the orientation of $\bm{A}_s$ relative to $\bm{A}_o$
    has an effect on the tolerance because the associated "thickness" of the ellipsoid changes.
    This information is captured by the plane containing vectors $\bm{A}_o$ and $\bm{A}_s$:
    \begin{align} \label{eq:10}
      (\bm{A}_s \times \bm{A}_o)^\top \bm{A} = 0
    \end{align}

    To find the constrained points of tangency, we can intersect the original ellipsoid with planes
    \ref{eq:9} and \ref{eq:10}. The line of intersection is such that
    \begin{align} \label{eq:11}
      \bm{A}(t) = (1 - \nicefrac{1}{c_1}) \bm{A}_o + t\bm{D}
    \end{align}

    \noindent where $c_1$ (equation \ref{eq:12}) is a location parameter projecting $\bm{A}_o$ onto plane \ref{eq:9}
    and $\bm{D}$ (equation \ref{eq:13}) is the direction perpendicular to planes \ref{eq:9} and \ref{eq:10}.

    \begin{align} \label{eq:12}
      \nonumber \bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} ((1 - \nicefrac{1}{c_1})\bm{A}_o - \bm{A}_o)  = -1\\
      \Rightarrow c_1 = \bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} \bm{A}_o = \|\bm{A}_o^\top, \bm{A}_o\|_{\hat{\bm{\Sigma}}^{-1}}^2
    \end{align}
    \begin{align} \label{eq:13}
      \nonumber\bm{D} &= (\bm{A}_o^\top \hat{\bm{\Sigma}}^{-1})^\top \times (\bm{A}_s \times \bm{A}_o)\\
        &= c_1\bm{A}_s - c_2\bm{A}_o
    \end{align}

@@

    Combining equations \ref{eq:7} and \ref{eq:11} produces a quadratic parameterized by the scalar $t$,
    from which we can obtain a quadratic equation for the points of tangency:

    \begin{align} \label{eq:14}
      \nonumber (t\bm{D} - \nicefrac{1}{c_1}\bm{A}_o)^\top \hat{\bm{\Sigma}}^{-1} (t\bm{D} - \nicefrac{1}{c_1}\bm{A}_o)  = 1\\
      ({\bm{D}^\top \hat{\bm{\Sigma}}^{-1} \bm{D}})t^2 - 
          \nicefrac{2}{c_1}(\bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} \bm{D})t + (\nicefrac{1}{c_1} - 1) = 0
    \end{align}

    Denote $c_3 = \bm{A}_s^\top \hat{\bm{\Sigma}}^{-1} \bm{A}_s = \|\bm{A}_s^\top, \bm{A}_s\|_{\hat{\bm{\Sigma}}^{-1}}^2$. Then by
    substitution we have
    ${\bm{D}^\top \hat{\bm{\Sigma}}^{-1} \bm{D}} = c_1(c_1c_3 - c_2^2)$ and ${\bm{A}_o^\top \hat{\bm{\Sigma}}^{-1} \bm{D}} = 0$.
    Solving then yields the necessary points (denoted $\bm{A}_b$) as
    \begin{align} \label{eq:15}
      \bm{A}_b = \left(1 - \frac{1}{c_1}\right)\bm{A}_o \pm
        \left(\frac{1}{c_1} \sqrt{\frac{c_1 - 1}{c_1c_3 - c_2^2}}\right)(c_1\bm{A}_s - c_2\bm{A}_o)
    \end{align}

    Equation \ref{eq:15} shows that for the points to be well-defined, the following conditions
    must hold relative to the inner-product space defined by $\hat{\bm{\Sigma}}^{-1}$:

@@

    Thus, given 2 nonparallel vectors $\bm{A}_o$ and $\bm{A}_s$ and positive definite covariance
    matrix $\hat{\bm{\Sigma}}$, we accept the simulation if $\hat{\bm{A}}_o^\top \hat{\bm{A}_s} \geq \text{tol} 
    = \min_b \hat{\bm{A}}_o^\top \hat{\bm{A}_b}$. The condition, more formally, is
    \begin{align} \label{eq:16}
      \max_b \hat{\bm{A}}_o^\top (\hat{\bm{A}}_b - \hat{\bm{A}}_s) \leq 0
    \end{align}

@@

    The vector $\bm{A} = (A^P, A^{SV}, A^{SH})^\top$ can be treated as a continuously differentiable nonlinear function of
    parameters $\bm{p} = (\psi, \delta, \lambda)^\top$, defined by equation \ref{eq:1}. We can thus apply
    directed search methods to minimize the following cost function:
    \begin{align} \label{eq:17}
      \mu(\bm{p}) = \hat{\bm{A}}_o^\top \hat{\bm{A}}
    \end{align}

    \textcolor{red}{Continue from here...}

    \begin{figure}[ht!]
      \centering
      \begin{tikzpicture}[every node/.style={circle, draw, minimum size=.8cm}]
        \node (P1) at (0,2) {$\psi$};
        \node (P2) at (0,0) {$\delta$};
        \node (P3) at (0,-2) {$\lambda$};

        \node (H1) at (2.6,3) {$s_R$};
        \node (H2) at (2.6,1.5) {$q_R$};
        \node (H3) at (2.6,0) {$p_R$};
        \node (H4) at (2.6,-1.5) {$q_L$};
        \node (H5) at (2.6,-3) {$p_L$};

        \node (A1) at (5,1.5) {\scriptsize $A^P$};
        \node (A2) at (5,0) {\scriptsize $A^{SV}$};
        \node (A3) at (5,-1.5) {\scriptsize $A^{SH}$};

        \node (M) at (7,0) {$\mu$};

        \foreach \p in {2,3}
          \foreach \h in {1,2,3,4,5}
            \draw (P\p) -- (H\h);

        % \foreach \h in {1,2,3,4,5}
        %   \foreach \a in {1,2,3}
        %     \draw[dashed] (H\h) -- (A\a);

        \foreach \a in {1,2,3}
          \draw (A\a) -- (M);

        % \draw[dashed] (P1) -- (H1);
        \draw (P1) -- (H2);
        \draw (P1) -- (H3);
        \draw (P1) -- (H4);
        \draw (P1) -- (H5);

        \draw (H1) -- (A3);
        \draw (H1) -- (A1);
        \draw (H1) -- (A2);
        \draw (H2) -- (A1);
        \draw (H2) -- (A2);
        \draw (H3) -- (A1);
        \draw (H3) -- (A2);
        \draw (H4) -- (A3);
        \draw (H5) -- (A3);
      \end{tikzpicture}
      \caption{Computational graph of forwayd model.}
      \label{fig:1}
    \end{figure}
